---
title: 'Brain Health QI preliminary data summary for baseline + intervention period 1 reporting period'
output:
  word_document:
    reference_docx: template2.docx
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  type: 1
  hco_name: "A"
editor_options:
  chunk_output_type: console
always_allow_html: true
---

<!-- Note: When running report use the following line -->
```{r, eval=F, include=FALSE, echo=FALSE}
#<!-- title: 'Brain Health QI preliminary data summary for baseline reporting period: `r paste0(hco.name2)`' -->
```

<!-- Note: I had Claude edit 'brain-health-data-report-custom-by-HCO-int1.Rmd ' for conciseness and the original script is now in archive/scripts. Check that file in case there are any errors from this script. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F, results = "hide",
                      fig.width = 7, fig.height = 9, knitr.kable.NA = '')
```

<!-- Note: this script was edited by Claude to make more concise and is based on 'brain-health-data-report-custom-by-HCO-int1.Rmd', currently in /archive/scripts/. If any problems, refer to the original script. -->

```{r libraries}

library(here)
library(gt); library(gtsummary)
library(tidyverse)
library(readxl); library(kableExtra); library(flextable)
library(legendry)
```


```{r load-data}
# note: data from  data-handling.Rmd

load( file="dat.RData") # df and df.wide.tot

# MODIFIED for Shiny app: type and hco.name come from params now

type = params$type
hco.name = params$hco_name

# MODIFIED for Shiny app: reports are anonymous. hco.name (the real org name,
# supplied by the app or the YAML default) is kept ONLY for internal logic
# below; the org label shown in facet strips and validation tables is
# neutralized here, and figure titles no longer append an org name.
df <- df |> mutate(hco = "HCO")
if (!is.null(df.wide.tot)) df.wide.tot <- df.wide.tot |> ungroup() |> mutate(hco = "HCO")

# --- INTERACTIVE USE (running chunks in RStudio, outside the Shiny app) ---
# params then takes the YAML defaults (type=1, hco_name="A").
# To work on the "B"/monthly version interactively, uncomment:
# type = 2; hco.name = "B"
 
```


```{r overall-data}

# data handling

age.strata     <- c("65-70", "71-80", "81+")
sex.strata     <- c("Female", "Male", "Unknown")
race.strata    <- c("Asian", "Native Hawaiian or Other Pacific Islander",
                    "American Indian or Alaska Native", "Black or African American",
                    "Black", "White", "Race other", "Other Race", "Race missing/unknown")
ethnicity.strata <- c("Hispanic", "Hispanic or Latino", "Not Hispanic or Latino",
                      "Non-Hispanic", "Mexican", "Ethnicity missing/unknown")
language.strata  <- c("English", "non-English", "Missing/unknown")

hco_labels <- c(A="A") 

# MODIFIED for Shiny app: hco now carries the real HCO name from the app
# (not "test"), so fall back to the name itself when it's not in hco_labels.

overall.dat <- df |>
  filter(strata %in% age.strata) |>
  mutate(hco2 = dplyr::coalesce(hco_labels[hco], hco)) |>
  group_by(measure, hco2, time.f) |>
  summarize(mean.pct = mean(pct), sum.n = sum(sample), .groups = "drop") |>
  mutate(label.pct = as.character(round(mean.pct, 1)))

overall.dat %>% as.data.frame()

```

```{r make-fig-fn}

make.fig <- function(strata.vec, strata.name, rhs.name, title.text, hco.list, sgroup) {

  formula.grid <- as.formula(paste0("measure ~", rhs.name))
  
  prep <- df |>
    filter(strata %in% strata.vec,  strata.group %in% sgroup, sample>10)

  tot <- prep |>
    group_by(measure, time.f, hco) |>
    summarize(tot.sample = sum(sample), .groups = "drop")

  pdat <- prep |>
    left_join(tot, by = c("measure", "time.f", "hco")) |>
    mutate(n.tot = paste0("n= ", prettyNum(tot.sample, big.mark = ",", trim = TRUE)))

  ggplot(pdat, aes(x = time.f, y = pct, group = strata, colour = strata)) +
    labs(title  = paste0("Rates by measure, time, ", strata.name, " strata", title.text),
         colour = paste0(strata.name, " groups"),
         x = "Time (quarter)", y = "Percent") +
    facet_grid(formula.grid, scales = "free") +
    geom_point(size = 3) +
    geom_line(linewidth = 1) +
    geom_text(aes(y = 0, label = n.tot), vjust = 1.5, hjust = 0.5, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0.2, 0.2))) +
    theme_bw(base_size = 15) +
    theme(axis.text.x   = element_text(angle = 45, hjust = 1),
          legend.position = "bottom",
          plot.title      = element_text(size = 10)) +
    guides(color = guide_legend(override.aes = aes(label = ""),
                                nrow=2, byrow=T))
}

```

```{r B-fig-fn}

# Helper for org B option-2 nested-axis figures
make.B.fig <- function(data, strata.col, strata.vec, facet_rhs, colour.lab, title.sfx) {
  ggplot(data |> filter(.data[[strata.col]] %in% strata.vec,
                         !.data[[strata.col]] %in% "Unknown"),
         aes(x = interaction(time, time.f2), y = pct,
             group = .data[[strata.col]], colour = .data[[strata.col]])) +
    labs(title  = paste0("org B: Percent outcome by measure, time, and ", title.sfx,
                         " strata using option 2 times"),
         colour = colour.lab,
         x = "Time (month, year, and quarter)", y = "Percent") +
    facet_grid(as.formula(paste0("measure ~ ", facet_rhs)), scales = "free") +
    geom_point() +
    geom_line(linewidth = 1) +
    geom_text(aes(y = 0, label = n.tot), vjust = 1.5, hjust = 0.5, size = 2) +
    scale_y_continuous(expand = expansion(mult = c(0.5, 0.2))) +
    guides(x     = "axis_nested",
           fill  = "legend_group",
           color = guide_legend(override.aes = aes(label = ""))) +
    theme_bw(base_size = 15) +
    theme(legend.position = "bottom", plot.title = element_text(size = 10))
}

# Pre-compute org B totals (reused across figures)
if (type == 2) {
  
  B.tot <- df |>
    group_by(measure, time.f, strata.group, time.f2, hco) |>
    summarize(tot.sample = sum(sample), .groups = "drop")

  all.dat2.prep <- df |>
    left_join(B.tot, by = c("measure", "time.f", "strata.group", "time.f2", "hco")) |>
    mutate(n.tot = paste0("n=\n", prettyNum(tot.sample, big.mark = ",", trim = TRUE)))
}

```

# Figures

<!-- ## Age strata -->

```{r fig-age}
make.fig(strata.vec = age.strata, strata.name = "age", sgroup = "Age",
         rhs.name = ".", title.text = "", hco.list = hco.name)
```



<!-- ## Sex strata -->

```{r fig-sex}
make.fig(strata.vec = sex.strata, strata.name = "sex", sgroup = "Sex",
         rhs.name = ".", title.text = "", hco.list = hco.name)
```

<!-- ## Race -->

```{r fig-race}
make.fig(strata.vec = race.strata, strata.name = "race", sgroup = "Race",
         rhs.name = "hco", title.text = "", hco.list = hco.name)
```


<!-- ## Ethnicity strata -->

```{r fig-ethnicity}
make.fig(strata.vec = ethnicity.strata, strata.name = "ethnicity", sgroup = "Ethnicity",
         rhs.name = "hco", title.text = "", hco.list = hco.name)
```

<!-- ## Preferred language strata -->

```{r fig-language}
if (type==1) # note: org B (type=2) does not have language.
  make.fig(strata.vec = language.strata, strata.name = "language", sgroup = "Language",
           rhs.name = "hco", title.text = "", hco.list = hco.name)
```

# Validation

## Data validation step 1

Check for any negative counts or counts with decimal points.

```{r val1, results='markup'}
v1 <- df |> filter(sample < 0 | outcome < 0)

if (nrow(v1) == 0) print("No items to check for validation step 1.") else kable(v1)
```

## Data validation step 2

Check for any counts in which the total sample (denominator) > outcome (numerator) count.

**Note**: only for HCOs who use 'option 1' data reporting template.

```{r val2, eval=T}

if(type==1){

  tots.sum.strata <- df |>
  group_by(hco, measure, strata.group, time) |>
  summarize(tot.sample = sum(sample), tot.outcome = sum(outcome), .groups = "drop")

# Language totals also cover Race and Ethnicity groups
# lang.rows <- df.wide.tot %>% filter(hco %in% hco.name) |> filter(strata.group == "Language")
# df.wide.tot. <- bind_rows(df.wide.tot %>% filter(hco %in% hco.name),
#                           mutate(lang.rows, strata.group = "Ethnicity"),
#                           mutate(lang.rows, strata.group = "Race"))
# df.wide.tot. %>% as.data.frame()


# note that the language strata also applies to the race and ethnicity groups. 
# create dups for those strata and add
# --------------------------------------------------------
df.wide.tot %>% as.data.frame() %>% tail(n=50)
df.wide.tot.dup = df.wide.tot[df.wide.tot$strata.group=="Language",]
# df.wide.tot.dup %>% as.data.frame()

df.wide.tot.eth = df.wide.tot.dup %>%
  mutate(strata.group="Ethnicity")
#df.wide.tot.eth

df.wide.tot.race = df.wide.tot.dup %>%
  mutate(strata.group="Race")

df.wide.tot. = rbind(df.wide.tot,
                     df.wide.tot.eth,
                     df.wide.tot.race)

df.wide.tot. %>% as.data.frame()

if(!hco.name %in% "org B"){
  
  both.tot.dat <- tots.sum.strata |>
    left_join(df.wide.tot., by = c("measure", "strata.group", "hco", "time")) |>
    mutate(diff.sample  = as.numeric(tot.sample)  - as.numeric(Sample),
           diff.outcome = as.numeric(tot.outcome) - as.numeric(Outcome))
  
  dat.2 <- both.tot.dat |> filter(abs(diff.sample) > 0 | abs(diff.outcome) > 0)
}

}

```


```{r val2-print, results='markup'}
if (type == 2) {
  print("No items to check for validation step 2.") 
} else if (nrow(dat.2) == 0) {
  print("No items to check for validation step 2.")
} else {
  kable(dat.2)
}
```

## Data validation step 3

Check if there are outcome (numerator) counts less than or equal to total sample (denominator).

```{r val3, results='markup'}
dat.3 <- df |> filter(sample < outcome)

if (nrow(dat.3) == 0) print("No items to check for validation step 3.") else kable(dat.3)

```

## Data validation step 4

Determine if there are rates with more than a 20 percent change from prior period.

```{r val4}
# Helper to add missing quarter columns
add_missing_quarters <- function(df) {
  for (q in c("Jan-Mar '25", "Apr-Jun '25", "Jul-Sep '25", "Oct-Dec '25", "Jan-Mar '26"))
    if (!q %in% names(df)) df[[q]] <- NA
  df
}

make_val4_flextable <- function(df, caption_text, header_cols, header_span) {
  df |>
    as_grouped_data(groups = c("measure", columns = NULL)) |>
    flextable() |>
    set_table_properties(layout = "autofit") |>
    set_caption(caption_text) |>
    bold(i = ~ !is.na(measure), bold = TRUE, part = "body") |>
    theme_zebra() |>
    # add_header_row(top = TRUE,
    #                values = c("", header_cols),
    #                colwidths = c(3, header_span)) |>
    align(i = 1, align = "center", part = "header")
}

if (type==2) {
  grouped.ft <- all.dat2. |>
    group_by(measure, strata.group, strata1, strata2) |>
    arrange(year, month) |>
    mutate(denom.change = ((sample - lag(sample)) / lag(sample)) * 100) |>
    filter(abs(denom.change) > 20, sample > 30, !(year == 2024 & month == 10)) |>
    dplyr::select(measure, strata.group, strata1, strata2, year, month, denom.change)
} else {
  grouped.ft <- df |>
    group_by(measure, strata.group, strata) |>
    arrange(time.f) |>
    mutate(pct.change   = round(((pct - lag(pct)) / lag(pct)) * 100, 0),
           pct          = round(pct, 0),
           denom.change = round(((sample - lag(sample)) / lag(sample)) * 100, 0)) |>
    filter(abs(denom.change) > 20, sample > 30, time.f != "Oct-Dec '24") |>
    dplyr::select(measure, strata.group, strata, time.f, denom.change)
}
```


### Table view

```{r val4-table, results='markup'}

caption_txt <- "Time periods with sample/denominator with more than a 20 percent change from prior period (and sample/denominator >30)"

if (type==2) {
  grouped.ft |>
    pivot_wider(names_from = c("year", "month"), values_from = denom.change) |>
    arrange(measure, strata.group, strata1, strata2) |>
    make_val4_flextable(
      caption_txt,
      "Percent change of sample/denominator from prior period", 12) |>
    set_header_labels(values = c("Measure", "Strata group", "Strata 1", "Strata 2",
                                 "11/2024", "12/2024", "1/2025", "2/2025", "3/2025",
                                 "4/2025", "5/2025", "6/2025", "7/2025", "8/2025", "9/2025",
                                 "10/2025", "11/2025", "12/2025", "1/2026", "2/2026", "3/2026"))
} else {
  grouped.ft |>
    pivot_wider(names_from = time.f, values_from = denom.change) |>
    arrange(measure, strata.group, strata) |>
    add_missing_quarters() |>
    dplyr::select(measure, strata.group, strata,
           `Jan-Mar '25`, `Apr-Jun '25`, `Jul-Sep '25`, `Oct-Dec '25`, `Jan-Mar '26`) |>
    make_val4_flextable(caption_txt,
                        "Percent change of sample/denominator from prior period", 5)
}

```


